A Linear Inside-Outside Algorithm for Correcting Sequencing Errors in Structured RNAs

### 2013-05-17 JCB/introduction_JCB.tex
Added a few 'todo' notes

Ribonucleic acids (RNAs) are found in every living organism, and exhibit a broad range of functions, ranging from catalyzing
chemical reactions, as the RNase P or the group II introns, hybridizing  messenger RNA to regulate gene expressions,
to ribosomal RNA (rRNA) synthesizing proteins. Those functions  require specific structures,  encoded in their nucleotide
sequence. Although the functions, and thus the structures, need to be preserved through various organisms, the sequences
can greatly differ from one organism to another. This sequence diversity coupled with the structural conservation is a fundamental
asset for evolutionary studies. To this end, algorithms to analyze the relationship between RNA mutants and structures are required.

>>>>

\TODOTous{Relire le papier et ajouter references apparues entre temps (ex: en preparant la presentation).}
Ribonucleic acids (RNAs) are found in every living organism, and exhibit a broad range of functions, ranging from catalyzing
chemical reactions, as the RNase P or the group II introns, hybridizing  messenger RNA to regulate gene expressions,
to ribosomal RNA (rRNA) synthesizing proteins. Those functions  require specific structures,  encoded in their nucleotide
sequence. Although the functions, and thus the structures, need to be preserved through various organisms, the sequences
can greatly differ from one organism to another. This sequence diversity coupled with the structural conservation is a fundamental
asset for evolutionary studies. To this end, algorithms to analyze the relationship between RNA mutants and structures are required.

### 2013-05-17 JCB/methods_JCB.tex
Added a few 'todo' notes

\subsubsection{Definitions.}
Let $B:=\left\{\Ab,\Cb,\Gb,\Ub\right\}$ be the set of nucleotides.
Given $s\in B^n$ an RNA sequence, let $s_i$ be the nucleotide at position $i$. Let $\Omega$ be a set of un-gapped RNA sequences of
length $n$, and $S$ a secondary structure without pseudoknots. 
Formally, if $(i,j)$ and $(k,l)$ are base pairs in $S$, there is no overlapping extremities
 $\{i,j\}\cap \{k,l\}=\varnothing$ and either the intersection is empty 
 ($[i,j]\cap[k,l]=\varnothing$) or one is included in the other ($[k,l]\subset[i,j]$ or 
 $[i,j]\subset[k,l])$.

@@

\TODOYann{}
\TODOVlad{Add illustration from RECOMB presentation}
In every case, the denominator is the sum of the partitions function of exactly $m$ mutations, 
for $m$ smaller or equal to our target $M$. The numerators are divided in the following three cases.
\begin{description}
\item[$S_i=-1$:] If the nucleotide at position $i$ is not paired, we are concerned by the weights
over all sequences which have at position $i$ nucleotide $x$, which is exactly the sum
of the values of $\Y{i-1,i+1}{m-\Kron_{x,s_i}}{x,x}$, for all $m$ between $0$ and $M$.
\item[$S_i=k>i$:] Since we need to respect the derivation of the secondary structure $S$, if 
position $i$ is paired, we must consider the two partition functions. The \emph{outside} of the 
base pair, and the \emph{inside}, for all possible values for the nucleotide at position $k$, and
all possible distribution of the mutant positions between the inside and outside of the base pair. We also add the term of isostericity for this specific base pair.
\item[$S_i=k<i$:] Same as above, but with position $i$ pairing with a lower position.
\end{description}

>>>>

\subsubsection{Definitions.}
Let $B:=\left\{\Ab,\Cb,\Gb,\Ub\right\}$ be the set of nucleotides.
Given $s\in B^n$ an RNA sequence, let $s_i$ be the nucleotide at position $i$. Let $\Omega$ be a set of un-gapped RNA sequences of
length $n$, and $S$ a secondary structure without pseudoknots. 
\TODOYann{Move above, define secondary structure and add fancy VARNA illustration}

Formally, if $(i,j)$ and $(k,l)$ are base pairs in $S$, there is no overlapping extremities
 $\{i,j\}\cap \{k,l\}=\varnothing$ and either the intersection is empty 
 ($[i,j]\cap[k,l]=\varnothing$) or one is included in the other ($[k,l]\subset[i,j]$ or 
 $[i,j]\subset[k,l])$.

@@

\TODOYann{Rephrase as tree constructs}
\TODOVlad{Add illustration of Inside/Outside combination (adapted from RECOMB presentation??)}
In every case, the denominator is the sum of the partitions function of exactly $m$ mutations, 
for $m$ smaller or equal to our target $M$. The numerators are divided in the following three cases.
\begin{description}
\item[$S_i=-1$:] If the nucleotide at position $i$ is not paired, we are concerned by the weights
over all sequences which have at position $i$ nucleotide $x$, which is exactly the sum
of the values of $\Y{i-1,i+1}{m-\Kron_{x,s_i}}{x,x}$, for all $m$ between $0$ and $M$.
\item[$S_i=k>i$:] Since we need to respect the derivation of the secondary structure $S$, if 
position $i$ is paired, we must consider the two partition functions. The \emph{outside} of the 
base pair, and the \emph{inside}, for all possible values for the nucleotide at position $k$, and
all possible distribution of the mutant positions between the inside and outside of the base pair. We also add the term of isostericity for this specific base pair.
\item[$S_i=k<i$:] Same as above, but with position $i$ pairing with a lower position.
\end{description}
